---
title: "Lab 7: Clustering in R"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---
```{r}
iris_data <- subset(datasets::iris, select=c(1:4))
target <- datasets::iris$Species

get_external_metrics <- function(computed, target) {
  true_positives <- 0
  true_negatives <- 0
  false_positives <- 0
  false_negatives <- 0 
  
  # for every pair of every item
  for (i in 1:length(target)) {
    for (j in 1:length(target)) {
      if (computed[i] == computed[j] && target[i] == target[j]) {
        true_positives <- true_positives + 1
      }
      else if (computed[i] != computed[j] && target[i] != target[j]) {
        true_negatives <- true_negatives + 1
      }
      else if (computed[i] == computed[j] && target[i] != target[j]) {
        false_positives <- false_positives + 1
      }
      else if (computed[i] != computed[j] && target[i] == target[j]) {
        false_negatives <- false_negatives + 1
      }
    }
  }
  
  precision = true_positives / (true_positives + false_positives)
  recall = true_positives / (true_positives + false_negatives)
  f_score = (2 * precision * recall) / (precision + recall)
  return(list(precision=precision, recall=recall, f_score=f_score))
}

print_metrics <- function(metrics) {
  cat("\nPrecision:", metrics$precision)
  cat("\nRecall:", metrics$recall)
  cat("\nF-score:",metrics$f_score)
}
```


### Question 3: Kmeans

3a,b. Run kmeans for selected k values and report the cluster size and F-score.
```{r}
for (k in c(2,3,4,5,7,9,11)) {
  clusters <- kmeans(iris_data, k)
  metrics <- get_external_metrics(clusters$cluster, target)
  
  cat("K value:", k)
  cat("\nCluster sizes:", clusters$size)
  print_metrics(metrics)
  cat("\n\n")
}
```

3c. The value of `k` that produces the highest F-score is 3, with an F-score of `0.824`.

3d. This is to be expected, since there are 3 original species in the dataset.


### Question 4: hclust

4a. Display the result of the `hclust` algorithm as a dendrogram.
```{r}
iris_dist <- dist(iris_data)
clusters <- hclust(iris_dist)
plot(clusters)
```

4b. Looking at the display, the optimal clustering threshold I see is 2 clusters, since the height of the 2 clusters is greatest (between 7 and 4). This is interesting because the actual number of clusters is 3, but 2 of those 3 are super close together and easily mistaken for each other. The height for 3 clusters is a super small gap between 3.5 and 4, which doesn't seem very optimal at all.

4c. The `kmeans` algorithm picked up on the fact that there are 3 clusters, but the `hclust` algorithm shows that there are only 2. The composition of the 3 clusters shows that 1 is separate, and the other 2 are very close together. `hclust` doesn't distinguish between those 2 close clusters.

  
### Question 5: dbscan

5a,b. Run dbscan for selected eps and report the cluster size and F-score.
```{r}
iris_matrix <- as.matrix(iris_data)
for (eps in c(0.2, 0.3, 0.4, 0.5, 0.6, 0.8, 1.0)) {
  clusters <- dbscan::dbscan(iris_matrix, eps)
  metrics <- get_external_metrics(clusters$cluster, target)
  
  cat("eps value:", eps)
  cat("\nCluster sizes:", as.data.frame(table(clusters$cluster))[,2])
  print_metrics(metrics)
  cat("\n\n")
}
```

5c. The value of `eps` that produces the highest F-score is `1.0`, with an F-score of `0.75`.

5d. This is an interesting result, since this value of `eps` actually only recognizes 2 clusters instead of the 3 that are actually present. Other `eps` values experimented with different numbers of clusters, and a few had 3 clusters but not a high enough F-score. A recall value of `1` means that there were no false negatives identified with `eps=1`, which is interesting and a bit worrying to me.


### Question 6: swiss


```{r}
swiss_data <- subset(datasets::swiss, select=c(1,2,3,4))
clusters <- kmeans(swiss_data, 2, nstart = 5)
protestant = names(which(clusters$cluster == 1))
catholic = names(which(clusters$cluster == 2))
```

A list of the Swiss cities that are predominantly Protestant:
```{r}
protestant
```

A list of the Swiss cities that are predominantly Catholic:
```{r}
catholic
```